Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-06-13 2017-Xudong-SMT/AVOCS-2018-RR-eric.tex
RR-version of the AVOCS paper.

\documentclass{lncs/llncs}
\usepackage{etex}

\usepackage[T1]{fontenc}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}

@@

%\newtheorem{theorem}{Theorem}[section]
\newtheorem{prop}[theorem]{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}
%\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{alg}[theorem]{Algorithm}
%\newtheorem{remark}[theorem]{Remark}
%\newtheorem{definition}[theorem]{Definition}
%\newtheorem{example}[theorem]{Example}
%\newtheorem{problem}[theorem]{Problem}
%\newtheorem{proof}[theorem]{Proof}

@@

\title{Using SMT engine to generate Symbolic Automata\thanks{This work was partially 
funded by the Associated Team FM4CPS
  between INRIA and ECNU, Shanghai}}
\author{ Xudong Qin\inst{1,2}  \ \ \  Simon Bliudze\inst{3} \ \ \ Eric Madelaine\inst{1}
  \ \ \  Min Zhang\inst{2}} 
\institute{Universit\'e C\^ote d'Azur, Inria, CNRS, I3S, 06902 Sophia Antipolis, France
        \and Shanghai Key Laboratory of Trustworthy Computing, ECNU, China
\and INRIA Lille -- Nord Europe, 40 avenue Halley,
59650 Villeneuve d'Ascq,
France}
\date{}                                           % Activate to display a given date or no date

                          % Activate to display a given date or no date

\begin{document}

\maketitle

@@

\begin{abstract}
Open pNets are used to model the behaviour of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called "architectures"
for BIP systems, that have been used in the framework of an ESA project and to specify the control software of
a nanosatellite at the EPFL Space Engineering Center. We use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties; we give 2 examples, a safety and a liveness
property.
\end{abstract}

@@

For other applications, such as the analysis of 
programming languages, there exist dedicated platforms using
external automatic theorem 
provers (ATP) or automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from a user model. 
The prover we use, which also happens to be Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories framework (SMT), is mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011} or others.

BIP
(Behaviour-Interaction-Priority)~\cite{bip} is a framework for the
component-based design of concurrent software and systems.  In
particular, the BIP tool-set comprises compilers for generating C/C++
code, executable by linking with one of the dedicated engines, which
implement the BIP operational semantics~\cite{BarBliu15-offer-scico}.  This
approach ensures that any property, shown to hold on a given BIP
model, will also hold by construction on the generated code.
BIP Architectures~\cite{AttieBBJS16-architectures-faoc} formalise
design patterns, which enforce global properties 
characterising the coordination among the components of the
system. They provide a compositional approach, ensuring
correctness by construction during the design of BIP models.
In~\cite{AttieBBJS16-architectures-faoc}, it was shown that
application of architectures is compositional \wrt safety properties,
\ie when several architectures are applied, each enforcing a safety
property, the resulting system satisfies their conjunction.

@@

\begin{example}[An open transition]
  This transition is generated by application of the vector \texttt{SV0},
  synchronizing the initial actions of pLTSs \texttt{Control} and
  \texttt{Timer} (see Fig.~\ref{schema:ArchFailure:pNet}), with an action of the hole \texttt{B} equal to
  \texttt{fail}. The local variable \texttt{t} of the \texttt{Timer}
  is assigned to its initial value \texttt{Max}.
 A full output of 
the use-case is provided in~\cite{Avocs-RR}.

@@

\paragraph{Structural Semantic Rules:}
The semantics of pNet in term of open automata has been defined in
\cite{henrio:Forte2016}, in the form of 2 structural rules, one for
pLTSs, one for pNet nodes. These rules are slightly improved, adding guards in the synchronisation vectors and syntax for universal quantifier in the guards (see~\cite{Avocs-RR}).

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule to build the OT\footnote{%
%
  We omit detailed presentation of this case for the sake of brevity.
%
}.
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Alg.~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

\paragraph{Matching.}
The \emph{Matching} method builds the OTs of a pNet node from those of
its subnets.
For each synchronisation vector and each possible
combination of behaviours of the subnets, as generated by the \emph
{Combining} method, it builds the corresponding open transition.
Here, we only detail the construction of the predicate.
From a synchronization vector
$sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$
and its guard $G_k$; a tuple of open transitions
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$, such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$; the hole behaviours $\mathit{Hole} = (b_j)^{j \in J}$; and a fresh variable $v$, representing the result action of the OT under construction, we build the predicate:
%
\begin{multline*}
  \mathit{MkPred}(sv, C, \overline{\mathit{hole}}) = 
  \\
  (\forall i\in I, a_i=a'_i) \land
%  (\forall i \not\in I, ot_i = -) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k.
\end{multline*}

@@

Our implementation submits satisfiability requests to
Z3 using its JAVA API. Here, for readability, we
show the Z3 code using its SMT-LIB input language.
Note that in the previous sections, the OTs were displayed in a
  simplified, human readable form. The input and output of our tool, and
  also the generated SMTlib fragments, are
  slightly more difficult to read, in 
particular because of structured names for the fresh variables generated
by the algorithm, allowing tracability of the result~\cite{Avocs-RR}.

@@

\emph{Translation of action algebra presentation.}
In~\cite{Avocs-RR}, we  define the translation of an algebra presentation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). We also formalize a condition of well-formedness of pNets ensures that the generated code is correct,
and will not raise runtime errors in the SMT engine. Note that the
\texttt{declare-datatypes} command comprises both the action
constructors from table \ref{table:BIPalgebra} and also the constant action
names from our example.
Additionally, we will include axiomatisation of any required functions
and predicates of the presentation data-types.

@@

\emph{Translation of open transitions.}
In~\cite{Avocs-RR} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
\item collect all variables in the transition, and declare them (using \texttt{declare-const})
\item check well-formedness and correct typing of expressions,
\item translate the predicate into a conjunction of assertions,
\item if present in the source state, translate the state-variable
  assignments into a disjunctive assertion.
\end{itemize}
%
This translation ensures that no runtime error will occur
in the SMT engine.

@@

To improve the readability of this figure, we used the following conventions:
we omit the transitions of the two pLTS, and the set of ``working''
holes; and we directly write the resulting action as first element of
each OT, rather than including it as an equality inside the
predicate.
%
%Notice, however that the loop transitions
%$\underline{\mathit{resume}}$ in states $\ostate{s0,t0}$ and
%$\ostate{s2,t0}$, corresponding, respectively, to open transitions
%$ot_2$ and $ot_9$ in Fig.~\ref{schema:ots} of
%App.~\ref{appendix:fullOutput}, involve $\OTvar{resume(false)}$ from
%the two pLTS, \ie only the hole, corresponding to the operand
%component of the Failure Monitor architecture, executes $\PortResume$.

@@

\emph{Scaling up.} One important motivation of this work is to
attack the complexity of verification of realistic systems by a
compositional and parametric approach. Still one may wonder if the price
for analysing our symbolic transitions will not make the approach too expensive
in term of computing time. We tried a
slighly bigger example, assembling 2 Failure controllers. In~\cite{Avocs-RR}, we show that Z3 can check the
satisfiablility of a 90K open transitions in a couple of minutes.

@@

%\newpage
%\appendix
%\section{Appendix}
The following appendices are included for the convenience of reviewers.
Our intention is to publish as a research report, before final publication, an
extended version containing all technical details, proofs, and more examples.

@@

\begin{theorem}[Finiteness of open-automata.]\\
Given an open pNet $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $pLTS_i^{i\in L}$ and holes $Hole_j^{j\in
  J}$, if the sets $L$ and $J$ are finite, if the synchronization vectors of all pNets 
  included in  $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ 
  are finite, and if
$\forall i \in L.\, finite{(states(pLTS_i))} \text{ and } pLTS_i$
has a finite number of state variables, then Algorithm 1 terminates
and produces an open automaton 
$\mathcal{T}$ with finitely many states and transitions.

@@

\begin{definition}[Fresh variable]\label{fresh-variable}
The format of \emph{fresh variable} (renaming) is defined as:
``$\emph{prefix : tree\ index\ :\ counter}$''.
\begin{itemize}
   \item[$\bullet$] $prefix$ identifies the kind of the
     variable. Current kinds are: sva\_<SV> (action variable in vector
     $SV$), ra (result action), hb\_<B> (behaviour of hole $B$).
   \item[$\bullet$] $tree\ index$ is the index of the node containing
     the variable in the tree-like structure of the pNet. 
   \item[$\bullet$] $counter$ is the current value of the corresponding counter.
\end{itemize}
\end{definition}

@@

\begin{table}\caption{\label{Table:predefinedSorts}\leftline{Algebra Presentation: predefined Sorts
  and Operators}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
                $\texttt{true},\ \texttt{false}$&
                $\land,\ \lor,\ \neg,\ \implies$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{FUN}$ &
                \\\specialrule{0em}{1pt}{1pt}
		Int 				&
                ${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
                $- \texttt{(unary)},\ +,\ -
                \texttt{(binary)},\ \times,\ \div, \texttt{etc.}$
                \\\specialrule{0em}{1pt}{1pt}
                \textsl{for any sort} & & $=,\ \ne$
		\\\hline
	\end{tabular}
\end{table}

\begin{table}\caption{\label{table:BIPoperators}\leftline{Additional
      operator for the BIP algebra}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
                \\\hline\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{FUN\_Action\_Bool}$ &
		\\\hline
	\end{tabular}
\end{table}

@@

\begin{table}\caption{\label{table:CCSalgebra}\leftline{Additional
      sorts and operators for the CCS algebra}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Channel, Data    			&
                $\emptyset$&
                $\emptyset$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{Emit(c,v)},\texttt{Receive(c,v)},\texttt{Tau}$ &
		\\\hline
	\end{tabular}
\end{table}

@@

\section{Translation to SMTlib input language}
\label{appendix:mapsmt}
The pNet elements, as defined above, can be full translated into
SMT-LIB language (\cite{BarFT-RR-17}), but there are a number of differences in the
structure of the models/languages, so the translation is not trivial.

@@

\subsubsection{Presentation Translation} %\TODO{Duplicate ? no,
                                %slightly more precise that in the
                                %paper body.}
We define here the translation of the algebra presentation, extended
with the constant operators collected from the pNet. It produces the
declaration of sorts (excepted Bool and Int) with their constructors
and selectors as \texttt{declare-datatypes}, and the declaration of
auxiliary operators as \texttt{declare-function} constructs.

@@

\begin{lstlisting}
Let Pres = <Sorts, Constrs, Ops> an extended presentation
(i.e. including the constants from the pnet),
- define MySorts = Sorts \ {Bool, Int}
- compute the strongly connected components in the graph
  of MySorts with respect to the relation is-using
- for each SCC in this graph, construct the definitions, using
\texttt{declare-datatypes} for sorts, their constants and
constructors, and \texttt{declare-function} for the other operators,
using the following translation function:

\end{lstlisting}

@@

		&&       		\\\specialrule{0em}{1pt}{1pt}
		\hline
\end{tabular}
%\end{table}

@@

\begin{example}
  For the CCS presentation in example~\ref{example:CCSpresentation}, we would get:
\begin{lstlisting}
  (declare-datatypes Data ())
  (declare-datatypes Channel ())
  (declare-datatypes ()
      ((Action (Emit (Chan_E Channel) (Value_E Data))
               (Receive (Chan_R Channel) (Value_R Data))
               Tau )))
  \end{lstlisting}
\end{example}

\begin{example}
  For our BIP use-case, we have built (see Fig.\ref{schema:smt-lib}):
  \begin{lstlisting}
  (declare-datatypes ()
      ((Action (fail) (resume) (timeout) (reset) (start) (tick) (ask)
    (FUN_Action_Bool (fst Action) (snd Bool))
    (Synchro (action Action)))))
  \end{lstlisting}
\end{example}

\subsubsection{Predicate translation}

>>>>

\documentclass[twoside]{article}
% \documentclass{lncs/llncs}
\usepackage{etex}
\usepackage[a4paper]{geometry}
\usepackage[latin1]{inputenc} % ou \usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc} % ou \usepackage[OT1]{fontenc}
\usepackage{graphicx}
% \usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{prop}[theorem]{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}
%\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{alg}[theorem]{Algorithm}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
%\newtheorem{problem}[theorem]{Problem}
%\newtheorem{proof}[theorem]{Proof}

@@

\usepackage{RR}
\RRNo{0000}
\RRdate{May 2018}
\RRauthor{% les auteurs
Xudong  \MakeUppercase{Qin}
\thanks{MoE Engineering Research Center for Software/Hardware Co-design Technology and Application, East China Normal University, 200062, Shanghai, China}
\and
Simon \MakeUppercase{Bliudze}
\thanks{INRIA Lille-Nord Europe, 40 avenue Halley, 59650 Villeneuve d'Asq, France}
\and
Eric \MakeUppercase{Madelaine}
\thanks[inria]{INRIA Sophia Antipolis M\'editerran\'ee, BP 93, 06902 Sophia Antipolis, France}
\and
Min  \MakeUppercase{Zhang}
\thanks{MoE Engineering Research Center for Software Hardware Co-design Technology and Application, East China Normal University, 200062, Shanghai, China}
}

%% Ceci apparait sur chaque page paire.
\authorhead{Qin \& Bliudze \& Madelaine \& Zhang}

\RRetitle{Using SMT engine to generate Symbolic Automata - Extended version}
\RRtitle{Utilisation d'un Moteur SMT pour g\'en\'erer des Automates Symboliques- Version \'etendue}
\titlehead{Using SMT engine to generate Symbolic Automata}

\RRnote{KAIROS is a common team between the I3S Lab (Univ. of Nice
  Sophia-Antipolis and CNRS) and INRIA Sophia Antipolis M\'editerran\'ee}

\RRabstract{Open pNets are used to model the behaviour of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows us to check properties of
such systems in a compositional manner. 
We implement an algorithm computing these semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called "architectures"
for BIP systems, that have been used in the framework of an ESA project and to specify the control software of
a nanosatellite at the EPFL Space Engineering Center. We use pNets to encode a BIP architecture
extended with explicit data, and compute its
open automaton semantics. This automaton may be used to prove
behavioural properties; we give 2 examples, a safety and a liveness
property.}

\RRresume{Les pNets ouverts sont utilisés pour modéliser le comportement des systèmes ouverts, synchrones ou asynchrones, exprimée dans divers calculs ou
langages de programmation. Ils sont dotés d'une sémantique opérationnelle symbolique en termes d'«Automata Ouverts». Cela nous permet de vérifier les propriétés de ces systèmes d'une manière compositionnelle.
Nous avons implémenté un algorithme calculant ces sémantiques,
en construisant des prédicats exprimant les conditions de synchronisation
entre les actions des composants du pNet. La vérification de tels
prédicats nécessite un raisonnement symbolique sur les logiques de premier ordre, mais également sur des données spécifiques à l'application. Nous utilisons le moteur SMT Z3 pour vérifier la
satisfiabilité des prédicats, et ne conserver dans l'automate ouvert que les
transitions satisfiables.

Nous illustrons notre approche par un exemple d'inspiration industrielle.
Pour cela nous partons d'«architectures» de systèmes BIP, qui ont été utilisés dans le cadre d'un projet de l'Agence Spatiale Européenne pour spécifier le logiciel de contrôle d'un nanosatellite au Centre d'ingénierie spatiale de l'EPFL. Nous utilisons les pNets pour encoder une architecture BIP
étendu avec des données explicites, et calculer sa
sémantique en termes d'automates ouverts. Cet automate peut être utilisé pour prouver des 
propriétés comportementales; nous donnons 2 exemples, une propriete de sureté et une de vivacité.}

\RRmotcle{S\'emantique comportementale symbolique, Analyse compositionelle de logiciels, R\'eseaux d'automates synchronis\'es}
\RRkeyword{Symbolic behavioral semantics, 
Compositional analysis of software systems, 
Networks of synchronized automata}
\RRprojet{Kairos}
%\RRprojetcommun{Scale}
\RCSophia % Sophia Antipolis M\'editerran\'ee

\begin{document}
\makeRR

@@

@@

For other applications, such as the analysis of 
programming languages, there exist dedicated platforms using
external automatic theorem 
provers (ATP) or automatic tactics from interactive theorem
provers (ITP), to perform symbolic reasoning, and for example to
discharge some subgoals in the proofs.
Tools like Rodin~\cite{deharbe2013,deharbe2014,abrial2007} have
already integrated several provers, like Z3, as modules for proving
the proof obligations generated from a user model. 
The prover we use, which also happens to be Z3, is developed by Microsoft Research
based on the satisfiability modulo 
theories framework (SMT), is mainly applied in extended static checking, test case
generation, and predicate abstraction.
In a similar way, there are several ATPs/ITPs we could consider to use for
result pruning and bisimulation checking in our algorithm, as an
alternative to Z3, such as CVC4~\cite{barrett:CAV2011},
Coq~\cite{armand:CPP2011},
Isabelle~\cite{blanchette:FroCoS2011} or others.

BIP(Behaviour-Interaction-Priority)~\cite{bip} is a framework for the
component-based design of concurrent software and systems.  In
particular, the BIP tool-set comprises compilers for generating C/C++
code, executable by linking with one of the dedicated engines, which
implement the BIP operational semantics~\cite{BarBliu15-offer-scico}.  This
approach ensures that any property, shown to hold on a given BIP
model, will also hold by construction on the generated code.
BIP Architectures~\cite{AttieBBJS16-architectures-faoc} formalise
design patterns, which enforce global properties 
characterising the coordination among the components of the
system. They provide a compositional approach, ensuring
correctness by construction during the design of BIP models.
In~\cite{AttieBBJS16-architectures-faoc}, it was shown that
application of architectures is compositional \wrt safety properties,
\ie when several architectures are applied, each enforcing a safety
property, the resulting system satisfies their conjunction.

@@

\begin{example}[An open transition]
  This transition is generated by application of the vector \texttt{SV0},
  synchronizing the initial actions of pLTSs \texttt{Control} and
  \texttt{Timer} (see Fig.~\ref{schema:ArchFailure:pNet}), with an action of the hole \texttt{B} equal to
  \texttt{fail}. The local variable \texttt{t} of the \texttt{Timer}
  is assigned to its initial value \texttt{Max}.
 A full output of 
the use-case is provided in appendice~\ref{appendix:fullOutput}.

@@

\paragraph{Structural Semantic Rules:}
The semantics of pNet in term of open automata has been defined in
\cite{henrio:Forte2016}, in the form of 2 structural rules, one for
pLTSs, one for pNet nodes.  For convenience we also include these rules in
App.~\ref{appendix:semRules}.

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule (see TR1 in
App.~\ref{appendix:semRules}) to build the OT.\footnote{%
%
  We omit detailed presentation of this case for the sake of brevity.
%
}
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Alg.~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

@@

\paragraph{Matching.}
The \emph{Matching} method builds the OTs of a pNet node from those of
its subnets (see rule Tr2 in App.~\ref{appendix:semRules}).
For each synchronisation vector and each possible
combination of behaviours of the subnets, as generated by the \emph
{Combining} method, it builds the corresponding open transition.
Here, we only detail the construction of the predicate.
From a synchronization vector
$sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$
and its guard $G_k$; a tuple of open transitions
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$, such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$; the hole behaviours $\mathit{Hole} = (b_j)^{j \in J}$; and a fresh variable $v$, representing the result action of the OT under construction, we build the predicate:

$$
  \mathit{MkPred}(sv, C, \overline{\mathit{hole}})=(\forall i\in I, a_i=a'_i) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k
$$

@@

Our implementation submits satisfiability requests to
Z3 using its JAVA API. Here, for readability, we
show the Z3 code using its SMT-LIB input language.
Note that in the previous sections, the OTs were displayed in a
  simplified, human readable form. The input and output of our tool, and
  also the generated SMTlib fragments, are
  slightly more difficult to read, in 
particular because of structured names for the fresh variables generated
by the algorithm, allowing tracability of the result (see
appendix~\ref{appendix:variables}).

@@

\emph{Translation of action algebra presentation.}
In Appendix~\ref{appendix:mapsmt}, we  define the condition of
well-formedness of an algebra presentation, and its translation into
SMTlib declarations (\texttt{declare-datatypes} and
\texttt{declare-fun}). It ensures that the generated code is correct,
and will not raise runtime errors in the SMT engine. Note that the
\texttt{declare-datatypes} command comprises both the action
constructors from table \ref{table:BIPalgebra} and also the constant action
names from our example.
Additionally, we will include axiomatisation of any required functions
and predicates of the presentation data-types.

@@

\emph{Translation of open transitions.}
In App.~\ref{appendix:mapsmt} we formally define all steps of the translation of each
open transition, including:
%
\begin{itemize}
\item collect all variables in the transition, and declare them (using \texttt{declare-const})
\item check well-formedness and correct typing of expressions,
\item translate the predicate into a conjunction of assertions,
\item if present in the source state, translate the state-variable
  assignments into a disjunctive assertion.
\end{itemize}
%
This translation ensures that no runtime error will occur
in the SMT engine.

@@

To improve the readability of this figure, we used the following conventions:
we omit the transitions of the two pLTS, and the set of ``working''
holes; and we directly write the resulting action as first element of
each OT, rather than including it as an equality inside the
predicate.
%
Notice, however that the loop transitions
$\underline{\mathit{resume}}$ in states $\ostate{s0,t0}$ and
$\ostate{s2,t0}$, corresponding, respectively, to open transitions
$ot_2$ and $ot_9$ in Fig.~\ref{schema:ots} of
App.~\ref{appendix:fullOutput}, involve $\OTvar{resume(false)}$ from
the two pLTS, \ie only the hole, corresponding to the operand
component of the Failure Monitor architecture, executes $\PortResume$.

@@

\emph{Scaling up.} One important motivation of this work is to
attack the complexity of verification of realistic systems by a
compositional and parametric approach. Still one may wonder if the price
for analysing our symbolic transitions will not make the approach too expensive
in term of computing time. We tried a
slighly bigger example, assembling 2 Failure controllers. In
appendix~\ref{appendix:scalingup}, we show that Z3 can check the
satisfiablility of a 90K open transitions in a couple of minutes.

@@

\newpage
\appendix
%\section{Appendix}
The following appendices are included for the convenience of reviewers.
Our intention is to publish as a research report, before final publication, an
extended version containing all technical details, proofs, and more examples.

@@

\begin{theorem}[Finiteness of open-automata.]
Given an open pNet $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ with leaves $pLTS_i^{i\in L}$ and holes $Hole_j^{j\in J}$, if the sets $L$ and $J$ are finite, if the synchronization vectors of all pNets 
  included in  $\mylangle \overline{\pNet},\overline{S}, \symb{SV}_k^{k\in K}\myrangle$ 
  are finite, and if
$\forall i \in L.\, finite{(states(pLTS_i))} \text{ and } pLTS_i$
has a finite number of state variables, then Algorithm 1 terminates
and produces an open automaton 
$\mathcal{T}$ with finitely many states and transitions.

@@

\begin{definition}[Fresh variable]\label{fresh-variable}
The format of \emph{fresh variable} (renaming) is defined as:
$$\emph{prefix : tree\ index\ :\ counter.}$$
\begin{itemize}
   \item[$\bullet$] $prefix$ identifies the kind of the
     variable. Current kinds are: sva\_<SV> (action variable in vector
     $SV$), ra (result action), hb\_<B> (behaviour of hole $B$).
   \item[$\bullet$] $tree\ index$ is the index of the node containing
     the variable in the tree-like structure of the pNet. 
   \item[$\bullet$] $counter$ is the current value of the corresponding counter.
\end{itemize}
\end{definition}

@@

\begin{table*}

\begin{minipage}{\textwidth}
	\caption{\label{Table:predefinedSorts}\leftline{Algebra Presentation: predefined Sorts and Operators}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Bool    			&
                $\texttt{true},\ \texttt{false}$&
                $\land,\ \lor,\ \neg,\ \implies$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{FUN}$ &
                \\\specialrule{0em}{1pt}{1pt}
		Int 				&
                ${0, \{i, -i\}_{i \in \texttt{Nat}}}$  &
                $- \texttt{(unary)},\ +,\ -
                \texttt{(binary)},\ \times,\ \div, \texttt{etc.}$
                \\\specialrule{0em}{1pt}{1pt}
                \textsl{for any sort} & & $=,\ \ne$
		\\\hline
	\end{tabular}
	\vspace{12ex}
\end{minipage}
\begin{minipage}{\textwidth}
\caption{\label{table:BIPoperators}\leftline{Additional operator for the BIP algebra}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
                \\\hline\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{FUN\_Action\_Bool}$ &
		\\\hline
	\end{tabular}
\vspace{12ex}
\end{minipage} 
\begin{minipage}{\textwidth}
	\caption{\label{table:CCSalgebra}\leftline{Additional sorts and operators for the CCS algebra}}
	\begin{tabular}{p{3cm}p{3cm}p{6cm}}
		\hline\specialrule{0em}{1pt}{1pt}
		Sort & Constructors & Auxiliary Operators
                \\\specialrule{0em}{1pt}{1pt}
		\hline\specialrule{0em}{3pt}{3pt}
		Channel, Data    			&
                $\emptyset$&
                $\emptyset$
                \\\specialrule{0em}{1pt}{1pt} 
		Action 			&  $\texttt{Emit(c,v)},\texttt{Receive(c,v)},\texttt{Tau}$ &
		\\\hline
	\end{tabular}
\end{minipage}

\end{table*}

@@

@@

\section{Translation to SMTlib input language}
\label{appendix:mapsmt}
The pNet elements, as defined above, can be full translated into
SMT-LIB language (\cite{BarFT-RR-17}), but there are a number of differences in the
structure of the models languages, so the translation is not trivial.

@@

\subsection{Presentation Translation} %\TODO{Duplicate ? no,
                                %slightly more precise that in the
                                %paper body.}
We define here the translation of the algebra presentation, extended
with the constant operators collected from the pNet. It produces the
declaration of sorts (excepted Bool and Int) with their constructors
and selectors as \texttt{declare-datatypes}, and the declaration of
auxiliary operators as \texttt{declare-function} constructs.

@@

%\begin{lstlisting}
Let Pres = <Sorts, Constrs, Ops> an extended presentation
(i.e. including the constants from the pnet),
- define MySorts = Sorts \ {Bool, Int}
- compute the strongly connected components in the graph
  of MySorts with respect to the relation is-using
- for each SCC in this graph, construct the definitions, using
\texttt{declare-datatypes} for sorts, their constants and
constructors, and \texttt{declare-function} for the other operators,
using the following translation function:
\\
%\end{lstlisting}

@@

		&&       		\\\specialrule{0em}{1pt}{1pt}
		\hline
		\\
\end{tabular}
%\end{table}

@@

\begin{table}
\begin{example}
  For the CCS presentation in example~\ref{example:CCSpresentation}, we would get:
\begin{lstlisting}
  (declare-datatypes Data ())
  (declare-datatypes Channel ())
  (declare-datatypes ()
      ((Action (Emit (Chan_E Channel) (Value_E Data))
               (Receive (Chan_R Channel) (Value_R Data))
               Tau )))
  \end{lstlisting}
\end{example}

\begin{example}
  For our BIP use-case, we have built (see Fig.\ref{schema:smt-lib}):
  \begin{lstlisting}
  (declare-datatypes ()
      ((Action (fail) (resume) (timeout) (reset) (start) (tick) (ask)
    (FUN_Action_Bool (fst Action) (snd Bool))
    (Synchro (action Action)))))
  \end{lstlisting}
 \end{example}
 \vspace{25ex}
\end{table}

\subsection{Predicate translation}
